---
layout: post
title: "Visualizing the 2015 Champions League Semifinal on Reddit"
excerpt: "Data visualization is a powerful tool for analyzing an event. In this post, I look at how the volume of posts on Reddit from soccer fans changes in response to games in the Champions League semifinal."
categories: [R, Viz]
comments: true
---

Compelling graphics help us understand the impact of real world actions and events. For example, plotting app downloads over time and highlighting the dates on which news articles about the app were published is a powerful first step in quantifying the success of a PR strategy. Visually seeing changes in metrics around events or actions generates discussion, theories, and insight.

Reddit is a rich corpus of online messageboard posts about an endless variety of topics, and Kaggle hosts a user friendly sqlite database of posts from May of 2015. During that month, four semifinalists remaining in the Champions League competition played for a spot in the final. In this post, I'll use the `ggplot2` package to visualize the volume of posts on Reddit during the competition.

```{r message=FALSE, warning=FALSE}
library(dplyr) # data manipulation
library(stringr) # string manipulation
library(knitr) # nice tables
library(ggplot2) # plotting
source("readData.R") # read and process from sqlite db
```

## Data

Reddit is organized by subreddits under which discussions about a certain topic occur. While there are many subreddits related to soccer, I'll focus solely on discussion that occurred under the [/r/soccer](http://www.reddit.com/r/soccer) subreddit. `dplyr` makes it very easy to read and filter data directly from sqlite (see [here](http://github.com/llefebure/reddit-champions-league/blob/master/readData.R) for the `readData` function definition).

```{r}
df <- readData()
```

Attached to each comment are a number of attributes. Of particular interest is "author\_flair\_text". Users in the /r/soccer subreddit use "flairs" to indicate a team logo that shows up next to their username when they post. I will assume that a user supports or is a fan of a team if they post using that team's flair. I will use the terms supporter and user in the context of a team synonymously to mean that they posted using that team's flair.

```{r}
names(df)
```

## Which teams do users support?

Barcelona, Bayern Munich, Juventus, and Real Madrid were the four semifinalists in 2015. These teams have varying levels of support as some are more internationally known than others (Manchester United for example would be expected to have more users than Juventus). Before looking at how the semifinal games affected the volume of posts, it is important to understand which teams are represented in the dataset. Below are the ten with the largest number of posts by supporters. Also shown is the number of unique users for each team.
```{r}
supporter_counts <- df %>% 
  filter(!is.na(author_flair_text)) %>%
  group_by(author_flair_text) %>%
  summarize(comments = n(), users = n_distinct(author))
kable(head(supporter_counts %>% 
             select(author_flair_text, comments, users) %>% 
             arrange(desc(comments)), n = 10), 
      align = "c", col.names = c("Team", "Comments", "Users"))
```

All four of the semifinalists appear here. If you look closely, you will see that some team's supporters are more active than others. Juventus, for example, has relatively few unique users, yet they sit very high in the table. To see this relationship more clearly, the plot below looks at the relationship between the number of comments and the number of unique users.

```{r}
fit_coef <- coef(lm(comments~users, supporter_counts)) # linear trendline
ggplot(supporter_counts, aes(x = users, y = comments)) +
  geom_point() +
  geom_text(aes(label = ifelse(users > 400, author_flair_text, "")), size = 4, hjust = -.05) +
  theme_light(base_size=16) + 
  labs(x = "# of Unique Users", y = "# of Comments") +
  xlim(c(0, 2900)) +
  ggtitle("Active Supporters") + 
  geom_abline(intercept = fit_coef[1], slope = fit_coef[2])
```

All of the semifinalists appear above the linear trend line in this scatterplot indicating more activity from users than that of an average team. This makes sense since we would expect users to post more when their team is playing in a major competition and in the news frequently.

Juventus in particular appears to lie the farthest from the trendline among the semifinalists. This could possibly be the result of extra hype surrounding the Italian side's first Champions League final appearance since 2003. Real Madrid, Bayern Munich, and Barcelona are all heavyweights, being crowned champions in 2014, 2013, and 2011 respectively. 

Another point of interest in this plot is the United States (USMNT) falling far below the trend line. The USMNT played no games in May 2015, so perhaps there was not much to talk about. Another contributing factor could be that more active Redditors switched flairs to their in season club side.

## Did users post more during the competition?

Each semifinal matchup was a two game series, and scores were added to determine which team advanced to the final. The first leg featured Real Madrid at Juventus on May 5 and Bayern Munich at Barcelona on May 6.

* Juventus 2 - 1 Real Madrid
* Barcelona 3 - 0 Bayern Munich

The second leg matches occurred a week later, with Barcelona at Bayern Munich on May 12, and Juventus at Real Madrid on May 13.

* Bayern Munich 3 - 2 Barcelona
* Real Madrid 1 - 1 Juventus

```{r}
overall_volume <- df %>% 
  group_by(created_date) %>% 
  summarize(comments = n())
ggplot(overall_volume, aes(x=created_date, y=comments)) +
  geom_line() +
  theme_light(base_size=16) +
  labs(x = "Date", y = "# of Posts", title = "Overall Post Volume") +
  geom_vline(aes(xintercept=as.numeric(as.Date("2015-05-05"))), linetype = 2) +
  geom_vline(aes(xintercept=as.numeric(as.Date("2015-05-06"))), linetype = 2) +
  geom_vline(aes(xintercept=as.numeric(as.Date("2015-05-12"))), linetype = 2) +
  geom_vline(aes(xintercept=as.numeric(as.Date("2015-05-13"))), linetype = 2)
```

Plotted above is the overall post volume during the month. As expected, there is a spike on gamedays. Games were played at 8:45 PM local time (UTC+2) and the time scale on the axis is UTC, so the count of comments on gameday includes pregame commentary as well as immediate postgame commentary. There is notably more volume for the May 6th and May 13th games likely because these two were arguably the better of the two in their respective weeks (the first was an unexpected blowout and the second was very high stakes). 

However, there seems to be an even larger spike later in the month. Was there an event bigger than the Champions League in the world of soccer during that time? On May 27th, [news](http://www.nytimes.com/interactive/2015/05/27/sports/soccer/the-fifa-officials-charged-with-corruption.html) broke of the indictment of several top FIFA officials in a major corruption scandal. Surely this generated much discussion over the ensuing days.

Next, I will narrow focus to only supporters of the semifinalists. As seen previously, the number of users for each team varies, so it makes sense to normalize the post volume instead of plotting the volume directly. I do this by looking at the number of comments per user as a proportion of the mean daily number of comments. More precisely, if **comm** is the number of comments by Barcelona supporters on a given day and **avg** is the average number of comments per day by Barcelona supporters, the normalized quantity is **(comm - avg)/avg**. Below, the normalized post volume is plotted against time.

```{r}
semis <- df %>%
  filter(author_flair_text %in% c("Barcelona", "Bayern München", "Juventus", "Real Madrid")) %>%
  group_by(created_date, author_flair_text) %>%
  summarize(comments = n()) %>%
  group_by(author_flair_text) %>%
  mutate(team_mean = mean(comments), dev = comments/team_mean - 1) %>%
  ungroup()
head(semis)
```

```{r}
ggplot(semis, aes(x=created_date, y=dev, group=author_flair_text, color = author_flair_text)) +
  geom_line() +
  theme_light(base_size=16) +
  labs(x = "Date", y = "Normalized Post Volume", title = "Normalized Volume of Posts") +
  scale_colour_discrete(name = "Team") + 
  geom_vline(aes(xintercept=as.numeric(as.Date("2015-05-05"))), linetype = 2) +
  geom_vline(aes(xintercept=as.numeric(as.Date("2015-05-06"))), linetype = 2) +
  geom_vline(aes(xintercept=as.numeric(as.Date("2015-05-12"))), linetype = 2) +
  geom_vline(aes(xintercept=as.numeric(as.Date("2015-05-13"))), linetype = 2)
```

There is a clear spike in activity on gamedays from both the winning and losing sides. The largest is from Bayern supporters on the day of their crippling defeat at Barcelona in the first leg. This game undoubtedly left them frustrated and eager to vent on Reddit. The defeat also meant that the next game in the series would be far less exciting because Bayern would have to win by at least three goals to have a chance at advancing -- an extremely unlikely result given the context. This could explain why the peak for the second game in that series is comparatively small. In contrast, the peaks for the other series, which was more even, are similar on both gamedays.

## Conclusion

The analysis presented here shows how activity on Reddit changed in response to the Champions League. It additionally revealed another event that was not the focus of this post -- news breaking about the FIFA corruption scandal. As a data scientist, I like to work on projects that connect to real events and decisions. Visualization is an important link in conveying that connection.